---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

flinty provides the scientist a way to test if their dataset contains multiple subpopulations. This is an important question if the scientist wishes to infer quantities of interest from the dataset, for instance population-level parameters, without overlooking any latent heterogeneities. 

- In population genetics, suppose genetic samples are collected from individuals, say in a country like Mexico whose population history records admixture events and endogamous mating by cultural ancestry. It is not surprising if a dataset contains individuals that belong to distinguishable genetic communities. Unfortunately, absent information like family pedigrees, we might not have external means of grouping the individuals  


Here, we illustrate one of the applications of  

```{r setup}
library(flintyR)
```

This is a test vignette. Basic application and population-genetic simulations will be included here.
